---
title: "Lab 8"
author: "Stephen R. Proulx"
date: "3/3/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rethinking)
library(bayesplot)
source("../helper.R")
```
# Bayesian Statistical Modeling Winter 2021
# Lab Exercise, Week 9
# 3/8/2023


## Load and process the data

```{r}
library(rethinking) 
data(reedfrogs)
d <- reedfrogs
dat <- list(
S = d$surv,
n = d$density,
tank = 1:nrow(d),
P = ifelse( d$pred=="no" , 0L , 1L ), 
ts = ifelse( d$size=="small" , 0L , 1L ),
dtreat = ifelse( d$density=="10" , 1L , ifelse( d$density=="25"  , 2L ,3L) )
) %>%view()
```


## Model 1

```{r}
m1.1 <- ulam( alist(
S ~ binomial( n , p ),
logit(p) <- abar+ a[tank] *sigma ,
a[tank] ~ normal(0,1 ), 
abar ~ normal( 0 , 1.5 ),
sigma ~ exponential( 1 )
), data=dat , chains=4 , cores=4 , iter=3000, log_lik=TRUE )
```

```{r}
precis(m1.1)
```

This is the simulation if we use the inferred tank-specific values.
```{r}
sim.m1 <- sim_df(m1.1, data=dat)
 

sim.m1.sum<- sim.m1 %>%
  group_by(tank)%>%
  summarise(
    sbar=mean(S)/mean(n),
    slow=quantile(S/n,0.1),
    shigh=quantile(S/n,0.9),
    nbar=mean(n),
    pred=mean(P)
  )%>%
  ungroup()

ggplot(sim.m1.sum, aes(x=tank, y=sbar))+
  geom_point(aes(shape=as.factor(pred),color=as.factor(nbar)))+
  geom_errorbar(aes(ymin=slow,ymax=shigh))+
  geom_point(data=dat,aes(x=tank,y=S/n),color="purple")

```




## Model 2
```{r}
# pred
m1.2 <- ulam( alist(
S ~ binomial( n , p ),
logit(p) <- abar+ bp*P+ a[tank] *sigma ,
bp~ normal(0,1),
a[tank] ~ normal(0, 1 ), 
abar ~ normal( 0 , 1.5 ),
sigma ~ exponential( 5 )
), data=dat , chains=4 , cores=4 , iter=3000, log_lik=TRUE )
```
```{r}
precis(m1.2)
```


What does WAIC say about our new model?
```{r}
compare(m1.1,m1.2)
```

This is the simulation if we use the inferred tank-specific values.
```{r}
sim.m2 <- sim_df(m1.2, data=dat)
 

sim.m2.sum<- sim.m2 %>%
  group_by(tank)%>%
  summarise(
    sbar=mean(S)/mean(n),
    slow=quantile(S/n,0.1),
    shigh=quantile(S/n,0.9),
    nbar=mean(n),
    pred=mean(P)
  )%>%
  ungroup()

ggplot(sim.m2.sum, aes(x=tank, y=sbar))+
  geom_point(aes(shape=as.factor(pred),color=as.factor(nbar)))+
  geom_errorbar(aes(ymin=slow,ymax=shigh))+
  geom_point(data=dat,aes(x=tank,y=S/n),color="purple")

```


## Model 3

```{r}
# size
m1.3 <- ulam( alist(
S ~ binomial( n , p ),
logit(p) <- abar+ bs*ts+ a[tank] *sigma ,
bs~ normal(0,1),
a[tank] ~ normal(0, 1 ), 
abar ~ normal( 0 , 1.5 ),
sigma ~ exponential( 5 )
), data=dat , chains=4 , cores=4 , iter=3000, log_lik=TRUE )
```


```{r}
precis(m1.3)
```

WAIC comparison of model 1.3 and 1.1
```{r}
compare(m1.1,m1.3)
```


```{r}
sim.m3 <- sim_df(m1.3, data=dat)
 

sim.m3.sum<- sim.m3 %>%
  group_by(tank)%>%
  summarise(
    sbar=mean(S)/mean(n),
    slow=quantile(S/n,0.1),
    shigh=quantile(S/n,0.9),
    nbar=mean(n),
    pred=mean(P)
  )%>%
  ungroup()

ggplot(sim.m3.sum, aes(x=tank, y=sbar))+
  geom_point(aes(shape=as.factor(pred),color=as.factor(nbar)))+
  geom_errorbar(aes(ymin=slow,ymax=shigh))+
  geom_point(data=dat,aes(x=tank,y=S/n),color="purple")

```

## Model 4
```{r}
# pred + size
m1.4 <- ulam( alist(
S ~ binomial( n , p ),
logit(p) <- abar+ bs*ts + bp*P+ a[tank] *sigma ,
bs~ normal(0,1),
bp~ normal(0,1),
a[tank] ~ normal(0, 1 ), 
abar ~ normal( 0 , 1.5 ),
sigma ~ exponential( 5 )
), data=dat , chains=4 , cores=4 , iter=3000, log_lik=TRUE )
```



```{r}
precis(m1.4)
```



WAIC comparison of model 1.2 and 1.3 and 1.4
```{r}
compare(m1.2,m1.3,m1.4)
```

```{r}
sim.m4 <- sim_df(m1.4, data=dat)
 

sim.m4.sum<- sim.m4 %>%
  group_by(tank)%>%
  summarise(
    sbar=mean(S)/mean(n),
    slow=quantile(S/n,0.1),
    shigh=quantile(S/n,0.9),
    nbar=mean(n),
    pred=mean(P)
  )%>%
  ungroup()

ggplot(sim.m4.sum, aes(x=tank, y=sbar))+
  geom_point(aes(shape=as.factor(pred),color=as.factor(nbar)))+
  geom_errorbar(aes(ymin=slow,ymax=shigh))+
  geom_point(data=dat,aes(x=tank,y=S/n),color="purple")

```

## Model 5


```{r}
# pred + size + interaction
m1.5 <- ulam( alist(
S ~ binomial( n , p ),
logit(p) <- abar+ bs*ts + bp*P + gamma*ts*P+ a[tank] *sigma ,
bs~ normal(0,1),
bp~ normal(0,1),
gamma ~ normal(0,1),
a[tank] ~ normal(0, 1 ), 
abar ~ normal( 0 , 1.5 ),
sigma ~ exponential( 5 )
), data=dat , chains=4 , cores=4 , iter=3000, log_lik=TRUE )
```



```{r}
precis(m1.5)
```



WAIC comparison of model 1.2 and 1.4 and 1.5
```{r}
compare(m1.2,m1.4,m1.5)
```
```{r}
sim.m5 <- sim_df(m1.5, data=dat)
 

sim.m5.sum<- sim.m5 %>%
  group_by(tank)%>%
  summarise(
    sbar=mean(S)/mean(n),
    slow=quantile(S/n,0.1),
    shigh=quantile(S/n,0.9),
    nbar=mean(n),
    pred=mean(P)
  )%>%
  ungroup()

ggplot(sim.m5.sum, aes(x=tank, y=sbar))+
  geom_point(aes(shape=as.factor(pred),color=as.factor(nbar)))+
  geom_errorbar(aes(ymin=slow,ymax=shigh))+
  geom_point(data=dat,aes(x=tank,y=S/n),color="purple")

```


## compare all the parameters from all the models.

Find the row for sigma in the table of coefficients. This will show us how much sigma has gone down in the multilevel models.
```{r}

coeftab( m1.1 , m1.2 , m1.3 , m1.4 , m1.5 )
 
```

